The \dealii Library, Version 9.5

### 2023-07-06 9.5/paper.tex
Last edits.

@@

solver.reinit_vector       = [ ](VectorType &x) {/*...*/};    
solver.residual            = [ ](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
solver.apply_jacobian      = [&](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.solve_with_jacobian = [&](const VectorType &src,
                                 VectorType       &dst,
                                 const double      tol) {/*...*/};

@@

\dealii{} has, for several releases already, used CUDA to offload some
operations onto GPUs. It has also had interfaces to CUDA-based linear
algebra libraries. Yet, the diversification of GPU platforms away from
a single vendor (Nvidia) has made it clear that we need a different
strategy to support what users want. As a consequence, \kokkos has
become a mandatory dependency of \dealii{} as part of the current
release; if it is not found on a given system during configuration
time, then the library will fall back on a copy of \kokkos stored in
the \texttt{bundled} directory in the same way as we already
interface with \boost.

@@

Vector and matrix classes of the \petsc wrappers have been extended with an additional constructor
that can wrap an already existing \petsc vector or matrix, respectively.
The \texttt{BlockVector} and
\texttt{BlockSparseMatrix} classes now internally use
\petsc nested objects, i.e. \texttt{VECNEST} and  \texttt{MATNEST} respectively.
We have added a new class \texttt{PETScWrappers::PreconditionShell} to support user-defined
preconditioning that can be simply customized as the following code
snippet shows:
\begin{c++}
PETScWrappers::PreconditionShell preconditioner(/*...*/);

preconditioner.vmult = [&](const VectorType &src,
                           VectorType       &dst) {/*...*/};

@@

The \petsc~\snes sub-package solves systems of nonlinear equations of the form $F(x) = 0$.
The interface to \snes has been modeled on the already existing interface
to the \kinsol solver from the \sundials package, and the nonlinear problem
is specified with a set of callbacks that are used by the solver.
%\begin{c++}
%PETScWrappers::NonlinearSolver<VectorType> solver(/*...*/);
%
%solver.residual            = [&](const VectorType &x,
%                                 VectorType       &F) {/*...*/};
%solver.setup_jacobian      = [&](const VectorType &x) {/*...*/};
%solver.solve_with_jacobian = [&](const VectorType &rhs,
%                                 VectorType       &sol) {/*...*/};
%solver.solve(x);
%\end{c++}
The default configuration is set up to use a Jacobian-free Newton-Krylov (JFNK)
approach~\cite{knoll2004jacobian}, using \texttt{solve\_with\_jacobian} as linear solver.
Numerous other solver configurations are possible and can be selected
programmatically using the constructor arguments,
or via the powerful command line customization of \petsc.
This includes, for example, Quasi-Newton methods, Anderson's acceleration, and
nonlinear preconditioning~\cite{brune2015composing}. The step-77
tutorial program has gained a section that discusses the use of these
new \snes interfaces.

The \petsc~\ts sub-package solves ODEs in explicit or implicit form \cite{abhyankar2018petsc}, i.e.:
\begin{eqnarray*}
%\frac{\partial u}{\partial t} = G(t,u) &\text{explicit},\\
%F(t,u,\frac{\partial u}{\partial t}) = 0 &\text{implicit}\\
\dot{u} = G(t,u) &\text{(explicit)},\\
F(t,u,\dot{u}) = 0 &\text{(implicit)}.\\
\end{eqnarray*}
The interface to \ts has been modeled on the already existing interfaces
to the \ida and \arkode solvers from the \sundials package. Specifically:
\begin{c++}
PETScWrappers::TimeDependentSolver<VectorType> solver(/*...*/);

@@

Secondly, we have added a wrapper to \texttt{NOX}, a
nonlinear solver library that is similar to both the \kinsol{} solver from
the \sundials{} package to which we already had interfaces, and also
to the \snes{} collections of functions from \petsc (see also
Section~\ref{sec:petsc}). 
%The interfaces to these three solvers are
%essentially the same and require setting function objects for
%computing the residual of the nonlinear problem, along with setting
%up, applying, and solving with the Jacobian matrix:
%\begin{c++}
%TrilinosWrappers::NOXSolver<VectorType> solver(/*...*/);
%
%solver.residual            = [ ](const VectorType &src,
%                                 VectorType       &dst) {/*...*/};
%solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
%solver.apply_jacobian      = [&](const VectorType &src,
%                                 VectorType       &dst) {/*...*/};
%solver.solve_with_jacobian = [&](const VectorType &src,
%                                 VectorType       &dst,
%                                 const double      tol) {/*...*/};
%
%solver.solve(solution);
%\end{c++}
%
We refer interested readers to our documentation for other more advanced
functions of this wrapper which are related to the differences in features of these libraries.
In particular, we have added the possibility to reuse the preconditioner between
nonlinear steps (also known as \textit{preconditioner lagging}), which is
unfortunately natively only  supported in the official \texttt{EPetra} implementations.

\todo[inline]{I don't understand this last sentence. Can you
  elaborate? At least in the \kinsol{} interfaces, you only update the
  preconditioner whenever setup\_jacobian() is called.}

@@

With the current release, \dealii{} has gained interfaces to several
external packages that are largely driven via \textit{callbacks} --
see for example the code example in Section~\ref{sec:trilinos} on how
\texttt{NOX} gains information about the residual vector, how it indicates to
user code that the Jacobian matrix needs to be rebuilt,
and how to solve linearized systems of equations as the sub-steps of
solving a nonlinear problem. Indeed, the interfaces to the nonlinear
solver \kinsol{} (as part of \sundials{}) and \snes{} (as part of \petsc, see
Section~\ref{sec:petsc} function in exactly the same way; the
ODE solver interfaces to \arkode{} and \ida{} (also part of \sundials{}) and \ts{}
(as part of \petsc{}) also use this style.

This substantially enlarged use of callbacks used by different backend
libraries raises the issue that each underlying package has its own
convention on how success or error codes of these callbacks should be
encoded. In the case of \petsc{}'s \snes{} and \ts{}, and for \trilinos{}'s \texttt{NOX},
success is indicated by a zero integer return value, whereas failure is
indicated by a nonzero return value. On the other hand, the \sundials{}
packages indicate success by a zero integer return value, a
recoverable failure with a positive value, and an irrecoverable failure
with a negative value. Neither of these conventions mesh well with C++
where error codes are generally indicated via exceptions.
It is conceivable that libraries we want to
interface with in the future use yet other conventions.

@@

The current release includes numerous updates to the matrix-free
infrastructure, including:
\begin{itemize}
\item In release 9.3, we enabled parallel $hp$-operations in the matrix-free infrastructure.
The infrastructure did not work properly in the case that certain cells
did not get any degrees of freedom due to the usage of \texttt{FE\_Nothing}. This has been
fixed now. Furthermore, \texttt{FE\_Nothing} now also works together
with discontinuous Lagrange elements (i.e., with the \texttt{FE\_DGQ} class). Due to the popularity of \texttt{FE\_Nothing} as a means to enable
or disable cells, we have introduced the new class
\texttt{ElementActivationAndDeactivationMatrixFree}, which wraps a \texttt{MatrixFree} object, only loops over all
active cells and optionally interprets faces between active and deactivated cells
as boundary faces. This functionality has enabled simulations in powder-bed-fusion additive
manufacturing in~\cite{proell2023highly}.
\item The matrix-free infrastructure allows interleaving cell loops with vector updates
  by providing \texttt{pre}/\texttt{post} functions that are run on index ranges.  The \dealii
  library uses this feature, e.g., to improve the performance of (preconditioned)
conjugate gradient solvers~\cite{kronbichler2022cg} as well as of relaxation and Chebyshev iterations (see Subsection~\ref{sec:lac}).
Up to release~9.3, the \texttt{pre}/\texttt{post} infrastructure was only supported for
continuous elements (cell loop); now, it also works for discontinuous
elements which also require face loops to assemble jump and penalty terms.
\item The operator \texttt{CellwiseInverseMassMatrix} now also efficiently
evaluates the inverse for coupling (dyadic) coefficients in the case of multiple
components:
\begin{align*}
\left(v_i, D_{ij} u_j  \right)_{\Omega^{(K)}}
\quad  1\le i,j \le c,
\end{align*}
with $c$ being the number of components and $D\in \mathbb{R}^{c\times c}$ a tensorial
coefficient. The algorithm relies on the construction of the element mass matrix,
\begin{align*}
M =  ( I_1 \otimes N^T) (D \otimes I_2 ) ( I_1 \otimes N),
\end{align*}
with $N$ being the tabulated values of shape functions at quadrature and
$I_1,I_2$ identity matrices associated to $c$ vector components and the
quadrature points, respectively.
The algorithms assumes a square $N$:
\begin{align*}
M^{-1} =  ( I_1 \otimes N^{-1}) (D^{-1} \otimes I_2 ) ( I_1 \otimes N^{-T}).
\end{align*}
For hypercube-shaped cells, $N^{-1}$ has an explicit representation
again in terms of tensor products; for example, in 3d it can be
expressed as $N^{-1} = N_{1D}^{-1} \otimes N_{1D}^{-1} \otimes
N_{1D}^{-1}$, allowing the use sum factorization \cite{kronbichler2016comparison}.
\end{itemize}

@@

\dealii{} currently uses C++14 as the language standard to which its
code base is written. It can also use the classes
\texttt{std::optional} and \texttt{std::variant} if the compiler
supports C++17, but falls back to implementations obtained via the
\boost library otherwise, and this is true also for a number of
individual functions that were introduced in C++17 or C++20.

@@

@@

While there are no new \dealii tutorial programs in this release, many
were extensively revised: Around 145 of the more than 2100 (non-merge)
commits that went into this release touched the tutorial, in some
cases adding substantial amounts of text.

@@

The 9.5 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_4_2_and_9_5_0.html}
     {around XXX incompatible changes};
\todo{Fix number before release}
see \cite{changes95}. Many of these
incompatibilities change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item
  The class \texttt{hp::DoFHandler} has been removed. The
  \texttt{DoFHandler} class now implements all $hp$-related
  functionality, as it has for the past two releases.

>>>>

\dealii{} has, for several releases already, used CUDA to offload some
operations onto GPUs. It has also had interfaces to CUDA-based linear
algebra libraries. Yet, the diversification of GPU platforms away from
a single vendor (Nvidia) has made it clear that we need a different
strategy to support what users want. As a consequence, \kokkos has
become a mandatory dependency of \dealii{} as part of the current
release; if it is not found on a given system during configuration
time, then the library will fall back on a copy of \kokkos stored in
the \texttt{bundled/} directory in the same way as we already
interface with \boost.

@@

solver.reinit_vector       = [&](VectorType &x) {/*...*/};    
solver.residual            = [&](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
solver.apply_jacobian      = [&](const VectorType &src,
                                 VectorType       &dst) {/*...*/};
solver.solve_with_jacobian = [&](const VectorType &src,
                                 VectorType       &dst,
                                 const double      tol) {/*...*/};

@@

@@

Vector and matrix classes of the \petsc wrappers have been extended with an additional constructor
that can wrap an already existing \petsc vector or matrix, respectively.
The \texttt{BlockVector} and
\texttt{BlockSparseMatrix} classes now internally use
\petsc nested objects, i.e. \texttt{VECNEST} and  \texttt{MATNEST} respectively.
We have added a new class \texttt{PETScWrappers::PreconditionShell} to support user-defined
preconditioning that can be simply customized as the following code
snippet shows:
\begin{c++}
PETScWrappers::PreconditionShell preconditioner(/*...*/);
preconditioner.vmult = [&](const VectorType &src,
                           VectorType       &dst) {/*...*/};

@@

The \petsc~\snes sub-package solves systems of nonlinear equations of the form $F(x) = 0$.
The interface to \snes in the class \texttt{PETScWrappers::NonlinearSolver} has been modeled on the already existing interface
to the \kinsol solver from the \sundials package: the nonlinear problem
is specified via a set of callbacks as shown above in Section~\ref{sec:nonlinear}.
%\begin{c++}
%PETScWrappers::NonlinearSolver<VectorType> solver(/*...*/);
%
%solver.residual            = [&](const VectorType &x,
%                                 VectorType       &F) {/*...*/};
%solver.setup_jacobian      = [&](const VectorType &x) {/*...*/};
%solver.solve_with_jacobian = [&](const VectorType &rhs,
%                                 VectorType       &sol) {/*...*/};
%solver.solve(x);
%\end{c++}
The default configuration is set up to use a Jacobian-free Newton-Krylov (JFNK)
approach~\cite{knoll2004jacobian}, using \texttt{solve\_with\_jacobian} as linear solver.
Numerous other solver configurations are possible and can be selected
programmatically using the constructor arguments,
or via the powerful command line customization of \petsc.
This includes, for example, Quasi-Newton methods, Anderson's acceleration, and
nonlinear preconditioning~\cite{brune2015composing}. The step-77
tutorial program has gained a section that discusses the use of these
new \snes interfaces.

The \petsc~\ts sub-package solves ODEs in explicit or implicit form \cite{abhyankar2018petsc}, i.e.:
\begin{eqnarray*}
%\frac{\partial u}{\partial t} = G(t,u) &\text{explicit},\\
%F(t,u,\frac{\partial u}{\partial t}) = 0 &\text{implicit}\\
\dot{u} = G(t,u) &\text{(explicit)},\\
F(t,u,\dot{u}) = 0 &\text{(implicit)}.
\end{eqnarray*}
The interface to \ts has been modeled on the already existing interfaces
to the \ida and \arkode solvers from the \sundials package. Specifically:
\begin{c++}
PETScWrappers::TimeDependentSolver<VectorType> solver(/*...*/);

@@

Secondly, we have added a wrapper to \texttt{NOX}, a
nonlinear solver library that is similar to both the \kinsol{} solver from
the \sundials{} package to which we already had interfaces, and also
to the \snes{} collections of functions from \petsc (see also
Section~\ref{sec:petsc}). 
%The interfaces to these three solvers are
%essentially the same and require setting function objects for
%computing the residual of the nonlinear problem, along with setting
%up, applying, and solving with the Jacobian matrix:
%\begin{c++}
%TrilinosWrappers::NOXSolver<VectorType> solver(/*...*/);
%
%solver.residual            = [ ](const VectorType &src,
%                                 VectorType       &dst) {/*...*/};
%solver.setup_jacobian      = [&](const VectorType &src) {/*...*/};
%solver.apply_jacobian      = [&](const VectorType &src,
%                                 VectorType       &dst) {/*...*/};
%solver.solve_with_jacobian = [&](const VectorType &src,
%                                 VectorType       &dst,
%                                 const double      tol) {/*...*/};
%
%solver.solve(solution);
%\end{c++}
%
As for the \snes{} interface above, the \texttt{NOX} interface is
drive by the callbacks shown in Section~\ref{sec:nonlinear}, and
implemented in the \texttt{TrilinosWrappers::NOXSolver} class.

@@

With the current release, \dealii{} has gained interfaces to several
external packages that are largely driven via \textit{callbacks} --
see for example the code examples for nonlinear or ODE solvers in the previous sections.

This substantially enlarged use of callbacks used by different backend
libraries raises the issue that each underlying package has its own
convention on how success or error codes of these callbacks should be
encoded. In the case of \petsc{}'s \snes{} and \ts{}, and for \trilinos{}'s \texttt{NOX},
success is indicated by a zero integer return value, whereas failure is
indicated by a nonzero return value. On the other hand, the \sundials{}
packages indicate success by a zero integer return value, a
recoverable failure with a positive value, and an irrecoverable failure
with a negative value. Newer PETSc versions can deal with recoverable
failures as well, in some cases by calling back into PETSc from a
callback to set an error flag, in others by setting the elements of a
returned vector to NaN.
None of these conventions mesh well with C++
where error codes are generally indicated via exceptions.
It is conceivable that libraries we want to
interface with in the future use yet other conventions.

@@

The current release includes numerous updates to the matrix-free
infrastructure, including:
\begin{itemize}
\item In release 9.3, we enabled parallel $hp$-operations in the matrix-free infrastructure.
The infrastructure did not work properly for cells
that do not have any degrees of freedom because they use \texttt{FE\_Nothing}. This has been
fixed now. Furthermore, \texttt{FE\_Nothing} now also works together
with discontinuous Lagrange elements (i.e., with the \texttt{FE\_DGQ} class). Due to the popularity of \texttt{FE\_Nothing} as a means to enable
or disable cells, we have introduced the new class
\texttt{ElementActivationAndDeactivationMatrixFree}, which wraps a \texttt{MatrixFree} object, only loops over all
active cells, and optionally interprets faces between active and deactivated cells
as boundary faces. This functionality has enabled simulations in powder-bed-fusion additive
manufacturing in~\cite{proell2023highly}.
\item The matrix-free infrastructure allows interleaving cell loops with vector updates
  by providing \texttt{pre}/\texttt{post} functions that are run on index ranges.  The \dealii
  library uses this feature, e.g., to improve the performance of (preconditioned)
conjugate gradient solvers~\cite{kronbichler2022cg} as well as of relaxation and Chebyshev iterations (see Subsection~\ref{sec:lac}).
Up to release~9.3, the \texttt{pre}/\texttt{post} infrastructure was only supported for
continuous elements (cell loop); now, it also works for discontinuous
elements which also require face loops to assemble jump and penalty terms.
\item The operator \texttt{CellwiseInverseMassMatrix} now also efficiently
evaluates the inverse for coupling (dyadic) coefficients in the case of multiple
components:
\begin{align*}
\left(v_i, D_{ij} u_j  \right)_{\Omega^{(K)}}
\quad  1\le i,j \le c,
\end{align*}
with $c$ being the number of components and $D\in \mathbb{R}^{c\times c}$ a tensorial
coefficient. The algorithm relies on the construction of the element mass matrix,
\begin{align*}
M =  ( I_1 \otimes N^T) (D \otimes I_2 ) ( I_1 \otimes N),
\end{align*}
with $N$ being the tabulated values of shape functions at quadrature and
$I_1,I_2$ identity matrices associated to $c$ vector components and the
quadrature points, respectively.
The algorithms assumes a square $N$:
\begin{align*}
M^{-1} =  ( I_1 \otimes N^{-1}) (D^{-1} \otimes I_2 ) ( I_1 \otimes N^{-T}).
\end{align*}
For hypercube-shaped cells, $N^{-1}$ has an explicit representation
again in terms of tensor products; for example, in 3d it can be
expressed as $N^{-1} = N_{1D}^{-1} \otimes N_{1D}^{-1} \otimes
N_{1D}^{-1}$, allowing the use sum factorization \cite{kronbichler2016comparison}.
\end{itemize}

@@

This version of \dealii{} uses C++14 as the language standard to which its
code base is written. It can also use the classes
\texttt{std::optional} and \texttt{std::variant} if the compiler
supports C++17, but falls back to implementations obtained via the
\boost library otherwise, and this is true also for a number of
individual functions that were introduced in C++17 or C++20.

@@

The next release of \dealii{} will build upon C++17.

@@

While there are no new \dealii tutorial programs in this release, many
were extensively revised: Around 145 of the more than 2200 (non-merge)
commits that went into this release touched the tutorial, in some
cases adding substantial amounts of text.

@@

The 9.5 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_4_2_and_9_5_0.html}
     {around 35 incompatible changes};
see \cite{changes95}. Many of these
incompatibilities change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item
  The class \texttt{hp::DoFHandler} has been removed. The
  \texttt{DoFHandler} class now implements all $hp$-related
  functionality, as it has for the past two releases.
